Nested Sampling with Multiple Objective Functions

### 2018-06-05 doc/ms.tex
Memorylessness omg

\subsection{Unique Properties of Nested Sampling}
What is special about the NS sequence of distributions?
Intuitively, the fact they are chosen adaptively provides a practical advantage
over annealing-based methods (and reduces the number of tuning parameters
substantially --- no `annealing schedule' is needed). While adaptive methods
for generating annealing schedules exist \citep{salomone2018unbiased}, these
are of no use in some phase change problems where a single temperature implies
a mixture of two or more phases.

Consider the Nested Sampling measure, proportional to the amount of time
NS spends in each region of the space (see \citet{polson2014vertical}
for further discussion of this concept, and a derivation of it from some
desiderata).
Since NS moves at a constant rate
in terms of $\log(X)$, this measure is proportional to
\begin{align}
m_{\rm NS} &\propto \frac{\pi(\xx)}{X(\xx)}.\label{eqn:measure}
\end{align}
The sequence of dead particles emitted by NS is then used to compute properties
of the posterior distribution, or other canonical distributions for temperatures
other than 1. Why does this work? Why does the NS measure of
Equation~\ref{eqn:ns_measure} allow the user to get a reasonable number of
samples for {\em any} canonical distribution?

We seek an algorithm for computing the partition function
$Z(\beta_1, \beta_2)$ for a range of values of the $\beta$s, from a single
run. To be a variant of Nested Sampling, the algorithm should satisfy
the following properties:
\begin{enumerate}
\item It should begin with $N$ points drawn from the prior $\pi(\xx)$.
\item It should seek to explore regions where the values of
$L_1(\xx)$ and $L_2(\xx)$ are higher than the prior $\pi(L_1, L_2)$
would typically imply.
\item The algorithm should consider a sequence of probability
distributions proportional to $\pi$, but restricted to smaller and smaller
domains for which the enclosed prior mass shrinks by a (roughly) constant and
known factor at each iteration.
\item The algorithm should be invariant to monotonic transformations of
$L_1$ and $L_2$, i.e. it should only depend on rankings of $L_1$ and $L_2$
values and not the values themselves.
\end{enumerate}
The standard Nested Sampling algorithm has these properties but for a
single objective function. The same is true of variants such as
Diffusive Nested Sampling \citep{dns, dnest4} and PolyChord
\citep{handley2015polychord}.

>>>>

\subsection{Unique Properties of Nested Sampling}
What is special about the NS sequence of distributions (the constrained priors)?
Intuitively, the fact they are chosen adaptively (i.e., the likelihood values
observed
at one iteration determine the likelihood threshold for the next constrained
distribution)
provides a practical advantage
over annealing-based methods (and reduces the number of tuning parameters
substantially --- no `annealing schedule' is needed). While adaptive methods
for generating annealing schedules exist \citep{salomone2018unbiased}, these
are of no use in some phase change problems where a single temperature implies
a mixture of two or more phases.

Another feature of NS is that its output can be used to reconstruct
any constrained prior. Consider the Nested Sampling measure, proportional
to the amount of time
NS spends in each region of the space (see \citet{polson2014vertical}
for further discussion of this concept, and a derivation of it from some
desiderata).
Since NS moves at a constant rate
in terms of $\log(X)$, this measure is proportional to
\begin{align}
m_{\rm NS} &\propto \frac{\pi(\xx)}{X(\xx)}.\label{eqn:measure}
\end{align}
and is improper if we imagine an NS run that does not terminate.
Consider using $m_{NS}$ as an importance sampling-type distribution to
estimate properties of a constrained prior
$p(\xx ; \ell) = \pi(\xx)\mathds{1}(L(\xx) > \ell)/X(\ell)$.
How useful $m_{\rm NS}$ is for determining properties of $p(\xx ; \ell)$
may be indicated by the KL-divergence from the former to the latter:
\begin{align}
D_{\rm KL}\left(p(\xx ; \ell) || m_{\rm NS}(\xx)\right)
    &= \int p(\xx ; \ell)
                \log \left(\frac{p(\xx ; \ell)}{m_{\rm NS}(\xx)}\right)
                \, d\xx \\
    &= \textnormal{const} + \int_{L(\xx) > \ell} \frac{\pi(\xx)}{X(\ell)}
                \log \left(\frac{X(L(\xx))}{X(\ell)}\right) \, d\xx
\end{align}
Rewriting the integral to be with respect to $X$, and taking advantage of the
fact that the implied prior for $X$ is uniform (since $X$ is the complementary
CDF of the likelihood values under $\pi$) gives:
\begin{align}
D_{\rm KL}\left(p(\xx ; \ell) || m_{\rm NS}(\xx)\right)
    &= \textnormal{const} + \int_0^{X(\ell)} \frac{1}{X(\ell)}
                \log \left(\frac{1}{X(\ell)}\right) \, dX
\end{align}

Remarkably, this does not depend on $\ell$. This means that every constrained
prior, no matter the value of its threshold $\ell$, is equally `distant'
from $m_{\rm NS}$. This can be seen intuitively using a standard
$\log L$-$\log X$ plot from NS. On these plots, the prior is a reflected
exponential distribution with scale 1, and the likelihood constraints
merely truncate the exponential distribution (by the memorylessness property
of the exponential, truncation is equivalent to shifting). The NS measure
is uniform in terms of $\log(X)$, so we have essentially shown that the KL
divergence from a wide uniform measure to a narrow shifted
exponential (of unit width) contained within it does not depend on the
position of the exponential.

Unfortunately, this property breaks down with more than
one objective function, and (as far as we know) the algorithm given in this
paper does not satisfy an analogous property.

\subsection{Desiderata}
While we could not see how to satisfy the property in the above subsection,
the algorithm of this paper does satisfy the following other NS properties:
\begin{enumerate}
\item It begins with $N$ points drawn from the prior $\pi(\xx)$.
\item It seeks to explore regions where the values of
$L_1(\xx)$ and $L_2(\xx)$ are higher than the prior $\pi(L_1, L_2)$
would typically imply.
\item It considers a sequence of probability
distributions proportional to $\pi$, but restricted to smaller and smaller
domains for which the enclosed prior mass shrinks by a (roughly) constant and
known factor at each iteration.
\item It is invariant to monotonic transformations of
$L_1$ and $L_2$, i.e. it should only depend on rankings of $L_1$ and $L_2$
values and not the values themselves.
\end{enumerate}
The standard Nested Sampling algorithm has these properties but for a
single objective function. The same is true of variants such as
Diffusive Nested Sampling \citep{dns, dnest4} and PolyChord
\citep{handley2015polychord}.
